---
title: "Austin Intakes Cleaning"
author: "Izzy Bedichek"
date: "2023-11-20"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(dplyr)
library(skimr)
library(tidyverse)
library(lubridate)
library(stringr)
library(knitr)
```

```{r originaldataframe}
knitr::opts_knit$set(root.dir = "../")
austin_intake_raw <- read.csv("../csv/Austin_Animal_Center_Intakes_20231106.csv")
head(austin_intake_raw)

austin_intake <- austin_intake_raw
```

 
```{r dataframe}
knitr::opts_knit$set(root.dir = "../")
lubridated_raw <- read.csv("../csv/lubridated.csv")

lubridated <- lubridated_raw
```

### Cleaning Messy Datasets with Stringr

```{r strings}
stringed <- lubridated %>% 
  mutate(Sex.upon.Intake = na_if(Sex.upon.Intake, "Unknown"),
         intake_neuter_spay_status = str_trim(as.character(str_extract_all(Sex.upon.Intake, "(.*) "))),
         intake_sex = str_trim(as.character(str_extract_all(Sex.upon.Intake, " (.*)")))) 

stringed %>% 
  select(intake_neuter_spay_status, intake_sex, Sex.upon.Intake, Animal.ID:year)
  
```

ref: [https://dplyr.tidyverse.org/reference/na_if.html]
### Cleaning Messy Datasets with Case_When

ref: https://dplyr.tidyverse.org/reference/case_when.html, https://gist.github.com/avallecam/bcdb8ca98b40e7276a7269b72e564d85,
https://stackoverflow.com/questions/63344455/if-else-statement-to-check-if-any-numbers-are-negative-in-r

```{r negativeage}
case_whend_negative_age <- stringed %>% 
  mutate(number_age_year = case_when(str_detect(Age.upon.Intake, "(.*) years") ~ 
                                       as.numeric(str_extract_all(Age.upon.Intake, "(.*) ")),
                                str_detect(Age.upon.Intake, "(.*) months") ~ 
                                  as.numeric(str_extract_all(Age.upon.Intake, "(.*) "))/12,
                                str_detect(Age.upon.Intake, "(.*) weeks") ~ 
                                  as.numeric(str_extract_all(Age.upon.Intake, "(.*) "))/(12*4),
                                .default = 999)) 

case_whend_negative_age %>% 
  select(Name, number_age_year, Age.upon.Intake, date) %>% 
  arrange(number_age_year) # looking for any 999s

```

I found negative ages, why? Here I looked for the names "Montopolis" and "Gorda" 
in austin_intake, in case my data cleaning created negative values

# making it easier to compare:
```{r compare}


austin_intake %>% 
  select(Age.upon.Intake, Name, DateTime) %>% 
  filter(Name == c("Montopolis", "Rascal", "Gorda")) %>% 
  arrange(Name, DateTime)

case_whend_negative_age %>% 
  select(number_age_year, Age.upon.Intake, Name, DateTime) %>% 
  filter(Name == c("Montopolis", "Rascal", "Gorda")) %>% 
  arrange(Name, DateTime)

```
ref: https://stackoverflow.com/questions/38036680/align-multiple-tables-side-by-side

I want to check where the ages got confused, I am going to check stringed

```{r verifyingage}

austin_intake %>% 
  select(Age.upon.Intake, Name, DateTime) %>% 
  filter(Name == c("Montopolis", "Rascal", "Gorda")) %>% 
  arrange(Name, DateTime)

stringed %>% 
  select(Age.upon.Intake, Name, DateTime) %>% 
  filter(Name == c("Montopolis", "Rascal", "Gorda")) %>% 
  arrange(Name, DateTime)

```

I checked a couple names and dates and it looks like austin_intake and stringed
don't have different/changed ages

```{r agefix}
case_whend_negative_age <- stringed %>% 
  mutate(number_age_year = 
           case_when(str_detect(Age.upon.Intake, "(.*) years") ~ 
                       as.numeric(str_extract_all(Age.upon.Intake, "(.*) ")),
                                str_detect(Age.upon.Intake, "(.*) months") ~ 
                       as.numeric(str_extract_all(Age.upon.Intake, "(.*) "))/12,
                                str_detect(Age.upon.Intake, "(.*) weeks") ~ 
                       as.numeric(str_extract_all(Age.upon.Intake, "(.*) "))/(12*4),
                                .default = 999)) 
```

I think that maybe people use the negative sign to mean "less than," as in
the animal is less than a year old?

```{r agehypothesis}

case_whend_negative_age %>% 
  select(number_age_year, Name, date, Age.upon.Intake) %>% 
  filter((number_age_year < 0) == TRUE) %>% 
  arrange(Name, date)

```

It looks like all of the negative values are -3, -2, or -1. 

There are also only 12 negative values. As such, since there is no way to 
confirm my idea, but it doesn't look like taking out the 12 negative values 
would impact the other 156,975 data points and my idea seems like a reasonable 
guess, I am going to replace all negative values with 999

```{r numberagetidy}
case_whend <- stringed %>% 
  mutate(number_age_year = case_when(str_detect(Age.upon.Intake, "(.*) years") ~ 
                                       as.numeric(str_extract_all(Age.upon.Intake, "(.*) ")),
                                     str_detect(Age.upon.Intake, "(.*) year") ~ 
                                       as.numeric(str_extract_all(Age.upon.Intake, "(.*) ")),
                                     str_detect(Age.upon.Intake, "(.*) months") ~ 
                                       as.numeric(str_extract_all(Age.upon.Intake, "(.*) "))/12,
                                     str_detect(Age.upon.Intake, "(.*) month") ~ 
                                       as.numeric(str_extract_all(Age.upon.Intake, "(.*) "))/12,
                                     str_detect(Age.upon.Intake, "(.*) weeks") ~ 
                                       as.numeric(str_extract_all(Age.upon.Intake, "(.*) "))/(12*4),
                                     str_detect(Age.upon.Intake, "(.*) week") ~ 
                                       as.numeric(str_extract_all(Age.upon.Intake, "(.*) "))/(12*4),
                                     str_detect(Age.upon.Intake, "(.*) days") ~ 
                                       as.numeric(0),
                                     str_detect(Age.upon.Intake, "(.*) day") ~ 
                                       as.numeric(0),
                                     .default = 999)) %>% 
  mutate(number_age_year = case_when(number_age_year >= 0 ~ number_age_year,
                                     number_age_year < 0 ~ 999,
                                     .default = number_age_year))

case_whend %>% 
  select(number_age_year, Name, date, Animal.Type, Age.upon.Intake) %>% 
  arrange(desc(number_age_year))

case_whend %>% 
  select(number_age_year, Name, date, Animal.Type, Age.upon.Intake) %>% 
  arrange(number_age_year)

```

 
```{r csv}
write.csv(case_whend, "../csv/case_whend.csv")
```
